\bf Model Visualization Techniques for a Social Network Model

### 2018-09-27 paper-child.Rnw
done with section 3 edits. plots still need work tho

%Every good \emph{data} analysis includes both numerical and visual summaries of the data, so why restrict \emph{model} analysis, description, and diagnostics to numerical summaries? 
\emph{Model visualization} complements traditional model diagnostic tools such as $R^2$ values and residual plots by enriching our understanding and interpretation of statistical models. Each of the three different meanings of the word ``model", the model \emph{family}, the model \emph{form}, and the \emph{fitted} model \citep[see][pg. 2]{modelvis}, can benefit from model visualization (model-vis). We apply the three principles of model-vis: view the model in the data space, visualize collections of models, and explore the process of fitting the model, to the models and data introduced in Section~\ref{sec:exdatamods}. Since we have already decided on our model family (SAOMs), we shift our focus to the fitted model and the model form. Specifically, we use model visualization to learn more about how the \emph{model form} affects the \emph{fitted model}. %In a modeling task, there are three different things called ``model" under consideration: . The latter is usally what first comes to mind when considering a model, where a specified model is fit, and parameter estimates are reported.% In the context of SAOMs, the \emph{fitted model} contains the form of the rate and objective functions, the estimated rate parameters, and the estimated objective function parameters. 
%The \textit{model form} describes the the model \textit{before} the fitting process, defining which parameters are in the model within the context of the larger model family. % In SAOMs, the \emph{model form} includes description of the rate and objective functions and the variables therein that describe how the network evolves over time. 
%Finally, the \textit{model family} is the broadest description of the model. This is the type of model that you wish to fit to the data, and is chosen based on the problem, data, and knowledge at hand. %For example, we chose to use a SAO \emph{model family} to model our network data over an exponential random graph model (ERGM) because we believe that actor-level variables effect network structure and formation, and we wanted to model the network changes over time.
%The model family, the model form, and the fitted model can each be visualized according to  %We begin by introducing the five models that we fit to our example data. 
%We use the data and models described in Section~\ref{sec:exdatamods} to guide our visual explorations of SAOMs under the three principles of model visualization. %We begin with novel tools and ideas to view a SAOM in the data space of a dynamic network. We then explore collections of the same models fit many times to the example data to see how the simulation processes in the \texttt{RSiena} ``black-box" affect the model fits. Finally, we look behind the scenes and into the individual steps of the continuous time Markov chain to learn more about how this ``hidden data" mechanism works and how it results in a fitted model.

The first way we hope to better understand SAOMs is by viewing the model in the data space. In \citeauthor{modelvis}, they chose define the \textit{data space} as ``the region over which we can reliably make inferences, usually a hypercube containing the data" \citep[p.~206]{modelvis}. But what does this definition mean for dynamic network models? We interpret the data space for our application as the combined data structures of the network: the node, edge, and time information.
%We want to make inferences on parameters in the model, but the multi-level data structure of networks makes the ``region of inference" difficult to define.
% %For dynamic social networks, there are a few different data ``spaces":
% \begin{enumerate}
% \item the actors and their corresponding covariates,
% \item the edges and their correpsonding variables that describe the relationships between the nodes, and
% \item the time, both the continuous unobserved time and the discrete observed time points, over which the network evolves
% \end{enumerate}
%The traditional node-link visualization uses one of many algorithms to layout the actors as points in 2D space, then draws segments connecting the points in 2D if there is an edge between two nodes, and draws nothing otherwise. The time aspect can be visualized by drawing each network observation in time and placing the observed timepoints side-by-side.

%The node, edge, and time data can be visualized together in various ways.% Because longitudinal network data consist of three different ``spaces" of data, viewing the model in the data space can depend on which aspect of the \textit{model} and the \textit{data} we are interested in viewing. %The SIENA software includes a method for modelling covariate change in the network
%Incorporating data covariates into the network structure allows us to assess whether the ties between nodes are affected by how nodes behave over time. In this instance, we would want to view predictions of edge presence over time as node variables change, visualizing all three data spaces at once. In addition, a SAOM can model node variable change over time, taking both the node and edge information into account. In this case, a plot of predicted covariate values over time would put the model into the time and node data space. %It is more intutive to view all of the data spaces simultaneously.

The node, edge, and time data can be visualized together in various ways, and one tool that can help is the \texttt{R} package \texttt{geomnet} \citep{geomnet}. The network's node or edge data can be mapped to different visual features in the node-link diagram.  The color, size, and shape of the points can be used to represent variables in the node data, while the color, linewidth, and linetype of the lines between points can be used to represent the edge variables. %In a social network, node data might be age, gender, and/or occupation of the person in the network, and edge data might be length of connection between two people, how the people first met (school, work, church, etc.), and/or how often they interact. 
To view temporal changes, we can view the network at different timepoints side-by-side to see the evolution. Pulling all of this information together with \texttt{geomnet} allows the entire data space to be viewed at once.

@@

To demonstrate, we use \texttt{geomnet} to visualize the connections in the $111^{th}$ United States Senate at two different points: when Hillary Clinton was in the senate, and after she left to become Secretary of State. Clinton was only in the $111^{th}$ senate from January 3-20, 2009, in the middle of her second term as the junior senator from New York. In that time, she authored two bills and was a cosponsor on 17s. There are many senators who cosponsored both of her bills, giving their edges to her a $WPC$ of 1. So with Clinton included in the node-link diagram in Figure~\ref{fig:funsenatevis}, the senate looks much more collaborative than it does without her. We then map our potential covariates to visual features: number of bills authored  maps to the size of the node, gender maps to the shape, and party to the color. In addition, we visualize the strength of the tie by mapping the $WPC$ value between the two senators to the linewidth of the edge. In this single visualization, we have viewed node information, edge information, and time.

@@

p1 <- ggplot(data = t1) +
  geom_net(aes(from_id = from, to_id = to, linewidth= linewidth),
           directed = T, curvature = .2, labelon=T, size = 5,
           arrow = arrow(angle = 20, type = "open", length = unit(.2, "cm")), ecolour = "black",
           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 4, arrowgap = .02,
           singletons = T) +
  labs(title = "Average: M1") + ThemeNet

p2 <- ggplot(data = filter(combinedavgs, cat == "1st Wave")) +
  geom_net(aes(from_id = from, to_id = to, linewidth= linewidth/3),
           directed = T, curvature = .2, labelon=T, size = 5,
           arrow = arrow(angle = 20, type = "open", length = unit(.2, "cm")), ecolour = "#969696",
           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 4, arrowgap = .03,
           singletons = T) +
   labs(title = "Wave 1") + ThemeNet +
    theme(plot.background = element_rect(color = 'black'))
p3 <- ggplot(data = filter(combinedavgs, cat == "2nd Wave")) +
  geom_net(aes(from_id = from, to_id = to, linewidth= linewidth/3),
           directed = T, curvature = .2, labelon=T, size = 5,
           arrow = arrow(angle = 20, type = "open", length = unit(.2, "cm")), ecolour = "#969696",
           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 4, arrowgap = .03,
           singletons = T) +
  labs(title = "Wave 2")  +  ThemeNet +
  theme(plot.background = element_rect(color = 'black'))
grid.arrange(p2, p1, p3, nrow = 1)
@

Another way to view the model in the data space is through simulation from the model. No one network simulated from a SAOM is going to look like the data or represent the model, so we might want to look at an \textit{expectation} from the network model. We frequently rely on expected values in data analyses, but network models, especially those as complex as SAOMs, lack an expected network value measure. We could talk about expected values of parameters, but the parameters can be hard to interpret. %Expected values of parameters can be computed, but if they cannot directly tell us anything about our dependent variable, they lose value. 
There is no defined expected value of a network model, though there are with most other statistical models. How then, can we arrive at an ``average" network? We answer this question through visualization.

%For network data, we view an ``average" network through a sort of summary network drawn using the traditional node-link diagram. 
In Figure~\ref{fig:summnetM1}, we show the \textit{expected network} of M1, created with 1,000 simulations of the second wave of the network from the model. To make this expected network, we first simulated 1,000 wave 2 and wave 3 observations of our small friendship example data from model M1, for which parameters had previously been estimated. We then aggregate the 1,000 simulated instances of wave 2, and by counting up the number of times each edge $i \rightarrow j$ appears in a network simulation. Then, we construct a single network for each time point with edge weights equal to the proportion of time the edge appears in our 1,000 simulations. This weighted network is shown in the middle panel of Figure~\ref{fig:summnetM1}, with edges present only in the average network if it appears in at least 51 of the 1,000 simulations. The edges that appear most frequently in the model are drawn with darker, thicker lines. On either side of the average network in Figure~\ref{fig:summnetM1}, we show the actual data, wave 1 on the left, and wave 2 on the right. We can see that the structure of the average network is much more similar to the first wave than to the second wave. However, the simulations are supposed to represent the second wave of data on the right of Figure~\ref{fig:summnetM1}. This is an indication that the simple model, M1, is doing a very poor job of capturing the change mechanism from the first to the second wave of observation. The average network can thus be used to help determine model goodness-of-fit. Because the the expected  network looks more like the first wave than the second wave, we can use the visualization in Figure~\ref{fig:summnetM1} as evidence of poor model fit.

@@

In Figure~\ref{fig:summary2}, we show \textit{difference networks} between the average network from M1 and the first and second waves of data. These networks are constructed by computing the difference between two adjacency matrices. The lack of model fit is very obvious here, with the green edges representing ties that only show up in the average network, and the red edges representing ties in the data that the model misses completely.  Again, the edgeweights demonstrate the frequency of the ties in the simulations from M1. On the left in Figure~\ref{fig:summary2}, we see the difference between the first wave of data and the average network. We only see green and black edges, meaning that the simulated networks consist primarily of edges in wave 1, plus some new edges that are not in the first wave of data. The additional green edges are all reciprocating black edges, meaning that the model predicts more reciprocity in ties than there actually is. On the right in Figure~\ref{fig:summary2}, we see the difference between the second wave of data and the average network. The model is simulating the second wave of data, so we would hope to see few differences between the average and wave2. However, there are six red edges, which are edges in wave 2 that do not appear at all in the average network, and there are several green edges that are in the average network, but not in the data. Although there are some edges that appear in both the average and wave 2, the majority of edges in the average are not in wave 2, and there are several missing edges. Thus, the model M1 is an incredibly poor fit, which we are able to see very easily by viewing the model in the data space in this way.

>>>>

%Every good \emph{data} analysis includes both numerical and visual summaries of the data, so why restrict \emph{model} analysis, description, and diagnostics to numerical summaries? 
\emph{Model visualization} complements traditional model diagnostic tools such as $R^2$ values and residual plots by enriching our understanding and interpretation of statistical models. Each of the three different meanings of the word ``model", the model \emph{family}, the model \emph{form}, and the \emph{fitted} model \citep[see][pg. 2]{modelvis}, can benefit from model visualization (model-vis). We apply the three principles of model-vis: view the model in the data space, visualize collections of models, and explore the process of fitting the model, to the models and data introduced in Section~\ref{sec:exdatamods}. Since we have already decided on our model family (SAOMs), we shift our focus to the fitted model and the model form. Specifically, we use model visualization to learn more about how the \emph{model form} affects the \emph{fitted model}.

First, we put models in the data space to assess the model fit. %In \citeauthor{modelvis}, they chose define the \textit{data space} as ``the region over which we can reliably make inferences, usually a hypercube containing the data" \citep[p.~206]{modelvis}. But what does this definition mean for dynamic network models?
We define the data space for SAOMs as the combined data structures of the network: the node, edge, and time information. We use the \texttt{R} package \texttt{geomnet} \citep{geomnet} to visualize the three data spaces together. The package lays out the network using node-link diagrams, and the network's node or edge data are then mapped to different visual features.  The color, size, and shape of the points can be used to represent variables in the node data, while the color, linewidth, and linetype of the lines between points can be used to represent the edge variables. To view temporal changes, we place the network at different timepoints side-by-side to see the evolution. By pulling all of this information together using \texttt{geomnet}, we can show the entire data space at once.

@@

To demonstrate the functionality of \texttt{geomnet}, we visualize the connections in the $111^{th}$ United States Senate at two different timepoints: when Hillary Clinton was in the senate, and after she left to become Secretary of State. Clinton was only in the $111^{th}$ senate from January 3-20, 2009, in the middle of her second term as the junior senator from New York. In that time, she authored two bills and was a cosponsor on 17. There are many senators who cosponsored both of her bills, giving their edges to her a $WPC$ of 1. So with Clinton included in the node-link diagram in Figure~\ref{fig:funsenatevis}, the senate looks much more collaborative than it does without her. We then map our potential covariates to visual features: number of bills authored  maps to the size of the node, gender maps to the shape, and party to the color. In addition, we visualize the strength of the tie by mapping the $WPC$ value between the two senators to the linewidth of the edge. By viewing node information, edge information, and time simultaneously, we have learned that Senator Clinton greatly affected the structure of collaboration within the senate.

@@

p1 <- ggplot(data = t1) +
  geom_net(aes(from_id = from, to_id = to, linewidth= linewidth),
           directed = T, curvature = .2, labelon=T, size = 4,
           arrow = arrow(angle = 20, type = "open", length = unit(.2, "cm")), ecolour = "black",
           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 3, arrowgap = .02,
           singletons = T) +
  labs(title = "M1 Expected Network") + ThemeNet + 
  theme(plot.background = element_rect(color = 'black'))

p2 <- ggplot(data = filter(combinedavgs, cat == "1st Wave")) +
  geom_net(aes(from_id = from, to_id = to, linewidth= linewidth/3),
           directed = T, curvature = .2, labelon=T, size = 4,
           arrow = arrow(angle = 20, type = "open", length = unit(.2, "cm")), ecolour = "#969696",
           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 3, arrowgap = .03,
           singletons = T) +
   labs(title = "Wave 1") + ThemeNet +
    theme(plot.background = element_rect(color = 'black'))
p3 <- ggplot(data = filter(combinedavgs, cat == "2nd Wave")) +
  geom_net(aes(from_id = from, to_id = to, linewidth= linewidth/3),
           directed = T, curvature = .2, labelon=T, size = 4,
           arrow = arrow(angle = 20, type = "open", length = unit(.2, "cm")), ecolour = "#969696",
           color = "#252525", hjust = .5, vjust = .5, labelcolour = 'white', fontsize = 3, arrowgap = .03,
           singletons = T) +
  labs(title = "Wave 2")  +  ThemeNet +
  theme(plot.background = element_rect(color = 'black'))
grid.arrange(p2, p1, p3, nrow = 1)
@

To view a SAOM in the data space, we simulate from the fitted model. No single network instance simulated from a SAOM will fully represent the model, so instead we look at an \textit{expectation} from the network model. We frequently rely on expected values in statistics, but network models, especially those as complex as SAOMs, lack an expected network measure. We could talk about expected values of parameters, but the parameters can be hard to interpret and cannot teach us much about the overall structure of the network. How then, can we arrive at an ``expected" network? We answer this question with visualization.
%Expected values of parameters can be computed, but if they cannot directly tell us anything about our dependent variable, they lose value. 
%There is no defined expected value of a network model, though expectations exist for most other statistical models.

%For network data, we view an ``average" network through a sort of summary network drawn using the traditional node-link diagram. 
%In Figure~\ref{fig:summnetM1}, we show the \textit{expected network} of wave 2 from model M1, created with 1,000 simulations from the model. 
To construct an expected network, we first simulated 1,000 wave 2 and wave 3 instances of our small friendship network from model M1, for which parameters had previously been estimated. We then counted the number of times each edge $i \rightarrow j$ appeared in one of the 1,000 network simulations. From these counts, we constructed one summary network for each time point, which has edge weights equal to the proportion of the simulated networks in which the edge appeared. Finally, we created the \textit{expected network} by including only the edges that appeared in more than 5\% of the simulations. The resulting expected network is in the middle panel of Figure~\ref{fig:summnetM1}. Edges with higher weights are darker, thicker lines. On either side of the average network in Figure~\ref{fig:summnetM1}, we show the data for comparison: wave 1 on the left, and wave 2 on the right. We can see that the structure of the average network is much more similar to wave 1 than to wave 2. In both wave 1 and the expected network, the edges form two distinct clusters of nodes: $\{2,3,4,5,6,12,13,15\}$ and $\{7,8,9,10,11,14\}$. The expected network, however, is supposed to represent the second wave of data, on the right in Figure~\ref{fig:summnetM1}. The expected network does not resemble wave 2, which has one small node cluster and on large node cluster, with four unconnected nodes. Because the expected network does not resemble wave 2, we take this as evidence that the simple model, M1, does not capture the change in network structure from the first to the second wave of data. Visualizing the expected network in this way places the model M1 in the data space, and has helped us assess model fit. % We can use the average network to help determine model goodness-of-fit. Because the the expected  network looks more like the first wave than the second wave, we can use the visualization in Figure~\ref{fig:summnetM1} as evidence of poor model fit.

@@

Another way to view a SAOM in the data space is with \textit{difference networks} between the expected network and the data, as shown in Figure~\ref{fig:summary2}. We created these networks by computing the difference between the adjacency matrices for the expected network and wave 1, and for wave 2 and the expected network. The lack of model fit becomes even more clear in Figure~\ref{fig:summary2}, with the XX NEED TO GO BACK AND REDO LINETYPES green XX edges representing ties that only show up in the expected network, and the red edges representing ties in the data that the model misses completely.  Again, the widths of the edges demonstrate the frequency of the ties in the simulations. On the left in Figure~\ref{fig:summary2}, we see the difference between the first wave of data and the expected network. We only see green and black edges, meaning that the simulated networks consist primarily of edges in wave 1, plus some new edges that are not in the first wave of data. The additional green edges are all reciprocating black edges, meaning that the model predicts more reciprocity in ties than there actually is. On the right in Figure~\ref{fig:summary2}, we see the difference between the second wave of data and the expected network. As the model is simulating the second wave of data, we would hope to see few differences between the expected network and wave 2. However, there are six red edges, which are edges in the data that do not appear at all in the expected network, and there are several green edges that are in the expected network, but not in the data. Although there are some edges that appear in both the expected network and wave 2, the majority of edges in the expected network are not in wave 2, and there are several missing edges. We conclude that the model M1 is a poor fit to our example data, which have seen very clearly by viewing the model in the data space.
